### 2022-10-27 dissertation/chapters/03-heuristics.tex
✨ (chapter3): Review Done

\noindent
Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train an \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

Chapter~\ref{chap:anns} briefly introduced the reader to the concept of \index{heuristic}\index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. This chapter presents more detailed background information on various different \index{heuristic}heuristics that have been used to train \acp{FFNN}. Broadly speaking, this chapter focuses on two different groups of \index{heuristic}heuristics, including classical gradient-based approaches and population-based \index{meta-heuristic}meta-heuristics. Each technique is presented and discussed in detail. Pseudo-code algorithms are provided for each technique and discussions follow on advantages, disadvantages, capabilities and limitations. The remainder of this chapter is structured as follows.

@@

      \item \textbf{Section~\ref{sec:heuristics:mh}} presents the reader with three different population-baed \index{meta-heuristic}meta-heuristics, including \acf{PSO}, \acf{DE} and \acfp{GA}.

@@

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

@@

Although gradient-based \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most widely used. Gradient-based heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acf{BP}.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function, relative to the weight at index $i$. The learning rate is an important hyper-parameter in gradient-based \index{heuristic}heuristics as it controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step as is shown in Equation~\eqref{eq:heuristics:gd:sgd} below.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.9\textwidth]{images/sgd.png}
      \caption{An illustration of \acf{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

@@

\begin{itemize}
      \item The appropriate value to use for the learning rate $\eta$ is difficult to determine and is often problem-specific. A learning rate that is too small causes premature exploitation, leading to slow and bad convergence. On the contrary, a learning rate that is too high may lead to bad learning outcomes as the heuristic keeps overshooting good minima.

      \item Learning rate schedules~\cite{ref:robbins:1951} can be introduced to dynamically change the learning rate throughout the training process, however, these schedules and their parameters have to be defined a priori and are thus problem specific~\cite{ref:darken:1992}.

@@

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that $m_{t}$ and $v_{t}$ presented above are estimates of the first moment (the mean) and the second moment (the uncentered variance) of the gradients respectively. ~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} mentions that because $m_{t}$ and $v_{t}$ are initialised to be vectors of 0's, they are biased towards 0. The aforementioned bias is especially prominent during the initial timesteps and/or when the decay rates $\beta_{1}$ and $\beta_{2}$ are small ($\beta_{1}$ and $\beta_{2}$ are close to 1). These biases need to be corrected. The bias-corrected first and second moment estimates are presented in Equations~\eqref{eq:heuristics:gd:adam_bc_first_moment} and~\eqref{eq:heuristics:gd:adam_bc_second_moment} below.

@@

\noindent
This section provided the reader with a collection of gradient-based \index{heuristic}heuristics. The next section presents the reader with \index{meta-heuristic}meta-heuristics.

\section{Meta-Heuristics}
\label{sec:heuristics:mh}

@@

This section aims to introduce the reader to the concept of \index{meta-heuristic}meta-heuristics, with focus on population-based meta-heuristics. Three well known \index{meta-heuristic}meta-heuristics, that have been shown to train \acs{FFNN}, are presented. These meta-heuristics include \acf{PSO}, \acf{DE} and \acfp{GA}.~\citeauthor{ref:carvalho:2006}~\cite{ref:carvalho:2006} compared various \acs{PSO} variants for training \acp{FFNN}.~\citeauthor{ref:espinal:2011}~\cite{ref:espinal:2011} compared \acs{DE} and \acs{PSO} when applied to \acs{FFNN} training and~\citeauthor{ref:gupta:1999}~\cite{ref:gupta:1999} compared \acs{BP} to a \acs{GA} for training \acp{FFNN}.

@@

The biggest difference between \index{meta-heuristic}meta-heuristics and gradient-bases heuristics is that
\index{meta-heuristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space~\cite{ref:blum:2003}.

\subsection{Particle Swarm Optimisation}
\label{sec:heuristics:mh:pso}

\Acf{PSO} is a stochastic population-based search algorithm based on the social behaviour of birds in a flock~\cite{ref:kennedy:1995}. By definition, the \acs{PSO} heuristic is nature-inspired.  \Acp{PSO} were first presented  by~\citeauthor{ref:kennedy:1995}\cite{ref:kennedy:1995}. Kennedy and Eberhart first applied \acp{PSO} to train of \acp{FFNN}~\cite{ref:eberhart:1995, ref:kennedy:1997}. The application of \acp{PSO} in the context of training \acs{FFNN} have been widely studied~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}. This section aims to provide the reader with the detail of \acs{PSO} implementation.

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \index{meta-heuristic}meta-heuristics.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyperdimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modeled such that they stochastically return to previously successful regions in the search space.

@@

      \item \textbf{Global Best Position}: Refers to a social memory construct, where the each particle has a reference to the best solution found in the particle's neighbourhood thus far. In the case of \text{gbest} \acs{PSO}, the global best position is the best position of the entire swarm. The personal best position is denoted $\hat{y}_{i}(t)$.
\end{itemize}

\noindent
During initialisation, particles are randomly places within the search space by sampling from a uniform distribution such that $x_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At timestep 0, the particle's initial position is set to be the particle's personal best solution such that $y_{i}(0) = x_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position} below.

@@

\noindent
In Equations~\eqref{eq:heuristics:pso:velocity} and~\eqref{eq:heuristics:pso:position} above, $i$ refers to the $i$-th particle in the swarm and $j$ refers to the $j$-th dimension of  the particle's position. A breakdown of the components introduced above is required. The velocity update step consists of:

@@

      \item \textbf{Social Component}: Denoted by the term $c_{2}r_{2+{j}}(t)[\hat{y}_{ij} - x_{ij}(t)]$. This component represents the particle's social  experience. It introduces an attractor to the swarm's best position so far. The social component is also stochastically scaled with random numbers $r_{2} \sim U(0,1)^J$, while also introducing the social acceleration coefficient $c_{2}$ that is used to control the influence of the social attractor.
\end{itemize}

\noindent
~\citeauthor{ref:vanwyk:2014}\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $V_{max}$ threshold, in essence, modeling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity_clamping}
      \begin{split}
            v_{ij}(t+1)=
            \begin{cases}
                  v'_{ij}(t+1) & \text{if } |v'_{ij}(t+1)| < V_{max,j}   \\
                  -V_{maxj}    & \text{if } v'_{ij}(t+1) \leq -V_{max,j} \\
                  V_{maxj}     & \text{if } v'_{ij}(t+1) \geq V_{max,j}
            \end{cases}
      \end{split}
\end{equation}

\noindent
$V_{max}$ then becomes another hyper-parameter that must be defined a priori. Appropriate values for $V_{max}$ may prevent swarm explosion, but also has an effect on the exploration and exploitation of the heuristic. If $V_{max}$ is small, particle update steps are small, resulting in exploitation~\cite{ref:eberhart:1996}. If $V_{max}$ is big, it allows for larger update steps, promoting more exploration.

@@

\subsection{Differential Evolution}
\label{sec:heuristics:mh:de}

This section aims to introduce the reader to the next population-based \acs{MH} called \acl{DE}. Similar to \acs{PSO}, \Ac{DE} is a stochastic population-based search strategy developed by Storm and Price~\cite{ref:price:2006} in 1995. \Ac{DE} shares a lot of similarities with other evolutionary \acs{MH} paradigms such as \acp{PSO} and \acp{GA}. However, \acs{DE} differs significantly in the sense that it makes use of distance and direction information from the current population to guide the search process~\cite{ref:engelbrecht:2007}. Originally, \acs{DE} was focused on multi-dimensional real-values optimisation problems, but unlike gradient-based heuristics, it does not require any gradient information. This means that \acs{DE} does not require the underlying optimisation problem to be differentiable, meaning that it can be applied to problems that are discrete, noisy and dynamic~\cite{ref:rocca:2011}.

Lots of research has been done on using \acs{DE} to train \acp{FFNN}. Some notable work include~\cite{ref:ilonen:2003}\cite{ref:slowik:2008}\cite{ref:mingguang:2009}. In these works, the authors often highlight the low computational complexity and simplicity of implementation for \acs{DE}.

Similar to other \acp{EA}, variation from one generation to the next is achieved through the application of crossover and mutation operators. Engelbrecht~\cite{ref:engelbrecht:2007} mentions that for other \acp{EA} if both these operators are used, crossover is applied first after which the generated offspring is mutated. For other \acp{EA} mutation step sizes are sampled from some probability distribution. \acs{DE} differs from these implementations in that

\begin{itemize}
      \item mutation is applied first to generate a \textit{trial vector}, which is then used within the crossover operator to produce one offspring and
      \item mutation step sizes are not sampled from prior known probability distributions.
\end{itemize}

In \acs{DE} mutation step sizes are influenced by the differences in positions of different entities in the current population. The positions of entities in the population provide valuable information about the fitness landscape. This is under the assumption that entities are initially uniformly placed in the search space. \acs{DE} aims to exploit this concept in order to find optimal solutions. There are three main components to the \acs{DE} heuristic. These include mutation, crossover and selection operators~\cite{ref:price:2006}. Each of these are presented briefly next.

\subsubsection{Mutation}
\label{sec:heuristics:mh:de:mutation}

The purpose of the mutation operator is to produce a trial vector for each entity in the current population by mutating a target vector with a weighted differential~\cite{ref:engelbrecht:2007}. This trial vector is used in the crossover operator (discussed next) to produce offspring. The mutation process then follows as such. For each each parent $x_{i}(t)$ generate a trial vector $u_{i}(t)$ as follows.

\begin{itemize}
      \item Select a target vector, $x_{i_{1}}(t)$ from the population the is not the same as the parent i.e. $i \neq i_{1}$.
      \item This is followed by randomly selecting two other individuals $x_{i_{2}}(t)$ and $x_{i_{3}}(t)$. Importantly, all of these entities must be unique such that $i \neq i_{1} \neq i_{2} \neq i_{3}$ and $i_{2}, i_{3} \sim U(1, N_{s}$ where $N_{s}$ is the size of the swarm.
      \item These individual entities are then used to calculate the trial vector by perturbing the target vector as presented in Equation~\eqref{eq:heuristics:de:mutation} below.
\end{itemize}

@@

In Equation~\eqref{eq:heuristics:de:mutation} $\beta \in (0, \infty)$ is the scale factor and controls the amplification of the differential variation~\cite{ref:engelbrecht:2007}. Note that in the above steps, the selection strategy to select the target vector is not specified. Selection is discussed in Section~\ref{sec:heuristics:mh:de:selection} below.

\subsubsection{Crossover}
\label{sec:heuristics:mh:de:crossover}

In the context of \acp{EA}, reproduction and recombination is done through the crossover operation. The same applies to \acs{DE}. The \acs{DE} crossover operator implements a discrete recombination of the trial vector, $u_{i}(t)$ (as was generated in Section~\ref{sec:heuristics:mh:de:mutation} above) and the parent vector $x_{i}(t)$ to produce new offspring $x'_{i}(t)$. The crossover operator is given in Equation~\eqref{eq:heuristics:de:crossover} below.

@@

In Equation~\eqref{eq:heuristics:de:crossover}, $x_{ij}(t)$ refers to the $j$-th element of the vector $x_{i}(t)$ and $\mathcal{J}$ refers to a set of crossover points or indices at which perturbation is done. Different techniques for determining the set, $\mathcal{J}$ has been proposed~\cite{ref:storn:1996}\cite{ref:storn:1997}. These include:

\begin{itemize}
      \item \textbf{Binomial crossover}: A crossover mask is generated by randomly selecting indices from the set of possible crossover points $\{1,2,\dots,N_{x}$ where $N_{x}$ is the problem dimension. This techique is presented in Algorithm~\ref{algo:heuristics:de:bin} below. In Algorithm~\ref{algo:heuristics:de:bin} $p_{r}$ is the crossover probability. The higher the value of $p_{r}$, the more points will be included in the set $\mathcal{J}$. A \textit{Bernoulli} distribution (presented in Chapter~\ref{chap:probability} can be used to generate the binomial crossover mask. Note that due to the probabilistic nature of this process, it is possible that no crossover points are selected. To counteract this situation, a randomly selected crossover point $j^{*}$ is included in the set $\mathcal{J}$ such that $\mathcal{J} \neq \emptyset$ where $\emptyset$ is the empty set.
      \item \textbf{Exponential crossover}:~\citeauthor{ref:engelbrecht:2007}\cite{ref:engelbrecht:2007} states that with exponential crossover a sequence of adjacent crossover points are selected start from some randomly selected crossover index. This means that the set of possible crossover points $\mathcal{J}$ is a circular array in indices. This technique does not require the selection of an additional crossover point $j^{*}$ as this technique includes at the very least one index, which is the starting index that is randomly selected. From this point, the next index is selected until $U(0,1) \geq p_{r}$ or $|\mathcal{J}| = N_{x}$. $p_{r}$ is the same crossover probability as mentioned above for binomial crossover. The implementation of exponential crossover is given in Algorithm~\ref{algo:heuristics:de:exp} below.
\end{itemize}

>>>>

\noindent
Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train a \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

Chapter~\ref{chap:anns} briefly introduced the reader to the concept of \index{heuristic}\index{heuristic}heuristics and \acfp{MH}s. This chapter presents more detailed background information on various different \index{heuristic}heuristics that have been used to train \acp{FFNN}. Broadly speaking, this chapter focuses on two different groups of \index{heuristic}heuristics, including classical gradient-based approaches and population-based \acp{MH}. Each technique is presented and discussed in detail. Pseudo-code algorithms are provided for each technique and discussions follow on advantages, disadvantages, capabilities and limitations. The remainder of this chapter is structured as follows.

@@

      \item \textbf{Section~\ref{sec:heuristics:mh}} presents the reader with three different population-baed \acp{MH}, including \acf{PSO}, \acf{DE} and \acfp{GA}.

@@

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}. Each of these are presented in detail in the following sections.

@@

Although gradient-based \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most widely used. Gradient-based heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \acp{MH} and other second-order derivative methods such as Newtons' method. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acf{GD} framework called \acf{BP}.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
A more detailed breakdown of the steps for \acs{BP}, as presented by ~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007}, is given in Appendix~\ref{app:backpropagation}.

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step as is shown in Equation~\eqref{eq:heuristics:gd:sgd} below.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.9\textwidth]{images/sgd.pdf}
      \caption{An illustration of \acf{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

@@

\begin{itemize}
      \item The appropriate value to use for the learning rate $\eta$, is difficult to determine and is often problem-specific. A learning rate that is too small causes premature exploitation, leading to slow and bad convergence. On the contrary, a learning rate that is too high may lead to bad learning outcomes as the heuristic keeps overshooting good minima.

      \item Learning rate schedules~\cite{ref:robbins:1951} can be introduced to dynamically change the learning rate throughout the training process, however, these schedules and their parameters have to be defined a priori and are often problem specific~\cite{ref:darken:1992}.

@@

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that $m_{t}$ and $v_{t}$ presented above are estimates of the first moment (the mean) and the second moment (the uncentered variance) of the gradients respectively. ~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} mentions that because $m_{t}$ and $v_{t}$ are initialised to be vectors of 0's, they are biased towards 0. The aforementioned bias is especially prominent during the initial timesteps and/or when the decay rates $\beta_{1}$ and $\beta_{2}$ are small ($\beta_{1}$ and $\beta_{2}$ are close to 1). The bias-corrected first and second moment estimates are presented in Equations~\eqref{eq:heuristics:gd:adam_bc_first_moment} and~\eqref{eq:heuristics:gd:adam_bc_second_moment} below.

@@

\noindent
This section provided the reader with a collection of gradient-based \index{heuristic}heuristics. The next section presents the reader with \acp{MH}.

\section{Meta-Heuristics}\label{sec:heuristics:mh}

@@

This section aims to introduce the reader to the concept of \acp{MH}, with focus on population-based meta-heuristics. Three well known \acp{MH}, that have been shown to train \acs{FFNN}, are presented. These meta-heuristics include \acf{PSO}, \acf{DE} and \acfp{GA}.~\citeauthor{ref:carvalho:2006}~\cite{ref:carvalho:2006} compared various \acs{PSO} variants for training \acp{FFNN}.~\citeauthor{ref:espinal:2011}~\cite{ref:espinal:2011} compared \acs{DE} and \acs{PSO} when applied to \acs{FFNN} training and~\citeauthor{ref:gupta:1999}~\cite{ref:gupta:1999} compared \acs{BP} to a \acs{GA} for training \acp{FFNN}.

@@

The biggest difference between \acp{MH} and gradient-bases heuristics is that
\acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space~\cite{ref:blum:2003}.

\subsection{Particle Swarm Optimisation}\label{sec:heuristics:mh:pso}

\Acf{PSO} is a stochastic population-based search algorithm based on the social behaviour of birds in a flock~\cite{ref:kennedy:1995}. By definition, the \acs{PSO} heuristic is nature-inspired.  \Acp{PSO} were first presented  by~\citeauthor{ref:kennedy:1995}\cite{ref:kennedy:1995}. Kennedy and Eberhart first applied \acp{PSO} to train of \acp{FFNN}~\cite{ref:eberhart:1995, ref:kennedy:1997}. The application of \acp{PSO} in the context of training \acs{FFNN} have been widely studied~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}. This section aims to provide the reader with the details of \acs{PSO} implementation.

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \acp{MH}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyperdimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modeled such that they stochastically return to previously successful regions in the search space.

@@

      \item \textbf{Global Best Position}: Refers to a social memory construct, where each particle has a reference to the best solution found in the particle's neighbourhood thus far. In the case of \text{gbest} \acs{PSO}, the global best position is the best position of the entire swarm. The personal best position is denoted $\hat{y}_{i}(t)$.
\end{itemize}

\noindent
During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $x_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At timestep 0, the particle's initial position is set to be the particle's personal best solution such that $y_{i}(0) = x_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position} below.

@@

\noindent
In Equations~\eqref{eq:heuristics:pso:velocity} and~\eqref{eq:heuristics:pso:position} above, $i$ refers to the $i$-th particle in the swarm and $j$ refers to the $j$-th dimension of  the particle's position. The velocity update step consists of:

@@

      \item \textbf{Social Component}: Denoted by the term $c_{2}r_{2_{j}}(t)[\hat{y}_{ij} - x_{ij}(t)]$. This component represents the particle's social  experience. It introduces an attractor to the swarm's best position so far. The social component is also stochastically scaled with random numbers $r_{2} \sim U(0,1)^J$, while also introducing the social acceleration coefficient $c_{2}$ that is used to control the influence of the social attractor.
\end{itemize}

\noindent
~\citeauthor{ref:vanwyk:2014}\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $V_{max}$ threshold, modeling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity_clamping}
      \begin{split}
            v_{ij}(t+1)=
            \begin{cases}
                  v'_{ij}(t+1) & \text{if } -V_{max,j} < v'_{ij}(t+1) < V_{max,j} \\
                  -V_{maxj}    & \text{if } v'_{ij}(t+1) \leq -V_{max,j}          \\
                  V_{maxj}     & \text{if } v'_{ij}(t+1) \geq V_{max,j}
            \end{cases}
      \end{split}
\end{equation}

\noindent
$V_{max}$ is another hyper-parameter that must be defined a priori. Appropriate values for $V_{max}$ may prevent swarm explosion, but also has an effect on the exploration and exploitation of the heuristic. If $V_{max}$ is small, particle update steps are small, resulting in exploitation~\cite{ref:eberhart:1996}. If $V_{max}$ is big, it allows for larger update steps, promoting more exploration.

@@

\subsection{Differential Evolution}\label{sec:heuristics:mh:de}

This section aims to introduce the reader to the next population-based \acs{MH} called \acf{DE}. Similar to \acs{PSO}, \acs{DE} is a stochastic population-based search strategy developed by Storm and Price~\cite{ref:price:2006} in 1995. \Ac{DE} shares a lot of similarities with other evolutionary \acs{MH} paradigms such as \acp{PSO} and \acp{GA}. However, \acs{DE} differs significantly in how distance and direction information from the current population is used to guide the search process~\cite{ref:engelbrecht:2007}. Originally, \acs{DE} was focused on multi-dimensional real-valued optimisation problems, but unlike gradient-based heuristics, it does not require any gradient information. \acs{DE} does not require the underlying optimisation problem to be differentiable and can thus be applied to problems that are discrete, noisy and dynamic~\cite{ref:rocca:2011}.

Lots of research has been done on using \acs{DE} to train \acp{FFNN}. Some notable work include~\cite{ref:ilonen:2003, ref:slowik:2008, ref:mingguang:2009}. In these works, the authors often highlight the low computational complexity and simplicity of implementation for \acs{DE}.

Similar to other \acp{EA}, variation from one generation to the next is achieved through the application of \index{crossover}crossover and \index{mutation}mutation operators. Engelbrecht~\cite{ref:engelbrecht:2007} mentions that for other \acp{EA}, if both \index{crossover}crossover and \index{mutation}mutation operators are used, \index{crossover}crossover is applied first, after which the generated offspring is mutated. Furthermore, other \acp{EA} sample \index{mutation}mutation step sizes from some probability distribution. \acs{DE} differs from the aforementioned \acp{EA} in two ways. Firstly, \index{mutation}mutation is applied first to generate a \textit{trial vector}, which is then used within the \index{crossover}crossover operator to produce one offspring. Secondly, \index{mutation}mutation step sizes are not sampled from prior known probability distributions.

In \acs{DE}, \index{mutation}mutation step sizes are influenced by the differences in positions of different entities in the current population. The positions of entities in the population provide valuable information about the fitness landscape, a concept \acs{DE} aims to exploit in order to find optimal solutions. There are three main components to the \acs{DE} heuristic. These include \index{mutation}mutation, \index{crossover}crossover and selection operators~\cite{ref:price:2006}. Each of these are presented in detail in the following sections.

\subsubsection{Mutation}\label{sec:heuristics:mh:de:mutation}

The purpose of the \index{mutation}mutation operator is to produce a trial vector for each entity in the current population by mutating a \textit{target} vector with a \textit{weighted differential}~\cite{ref:engelbrecht:2007}. The trial vector is used in the \index{crossover}crossover operator to produce offspring. The \index{mutation}mutation process then follows as such. For each each parent $x_{i}(t)$ generate a trial vector $u_{i}(t)$ as follows.

\begin{itemize}
      \item Select a target vector, $x_{i_{1}}(t)$ from the population that is not the same as the parent i.e. $i \neq i_{1}$. Various different selection strategies can be used to select the target vector.

      \item Randomly select two other individuals $x_{i_{2}}(t)$ and $x_{i_{3}}(t)$. Importantly, all of these entities must be unique such that $i \neq i_{1} \neq i_{2} \neq i_{3}$ and $i_{2}, i_{3} \sim U(1, N)$ where $N$ is the size of the swarm/population.

      \item Selected individual entities are then used to calculate the trial vector by perturbing the target vector as presented in Equation~\eqref{eq:heuristics:de:mutation} below.
\end{itemize}

@@

\noindent
In Equation~\eqref{eq:heuristics:de:mutation} $\beta \in (0, \infty)$ is the scale factor and controls the amplification of the differential variation~\cite{ref:engelbrecht:2007}.

\subsubsection{Crossover}\label{sec:heuristics:mh:de:crossover}

In the context of \acp{EA}, reproduction and recombination is done through the \index{crossover}crossover operation. The same applies to \acs{DE}. The \acs{DE} \index{crossover}crossover operator implements a discrete recombination of the trial vector, $u_{i}(t)$, as was generated in Equation~\eqref{eq:heuristics:de:mutation} above, and the parent vector $x_{i}(t)$ to produce new offspring $x'_{i}(t)$. The \index{crossover}crossover operator is given in Equation~\eqref{eq:heuristics:de:crossover} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:de:crossover}, $x_{ij}(t)$ refers to the $j$-th element of the vector $x_{i}(t)$ and $\mathcal{J}$ refers to a set of \index{crossover}crossover points or indices at which perturbation is done. Different techniques for determining the set $\mathcal{J}$, has been proposed~\cite{ref:storn:1996, ref:storn:1997}. These include:

\begin{itemize}
      \item \textbf{Binomial \index{crossover}crossover}: A \index{crossover}crossover mask is generated by randomly selecting indices from the set of possible \index{crossover}crossover points $\{1,2,\dots,J\}$ where $J$ is the problem dimension. Binomial \index{crossover}crossover is presented in Algorithm~\ref{algo:heuristics:de:bin}. In Algorithm~\ref{algo:heuristics:de:bin}, $p_{r}$ is the \index{crossover}crossover probability. The higher the value of $p_{r}$, the more points will be included in the set $\mathcal{J}$. A \textit{Bernoulli} distribution can be used to generate the binomial \index{crossover}crossover mask. Note that due to the probabilistic nature of this process, it is possible that no \index{crossover}crossover points are selected. To counteract this situation, a randomly selected \index{crossover}crossover point $j^{*}$ is included in the set $\mathcal{J}$ such that $\mathcal{J} \neq \emptyset$ where $\emptyset$ is the empty set.
